---
title: "R Notebook"
output: html_notebook
---

```{r imports}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
#install.packages("fredr")
library(fredr)
#install.packages("DescTools")
library(DescTools)
#install.packages("pracma")
library(pracma)
#install.packages("seastests")
library(seastests)
#install.packages(c("seasonal", "x13binary"))
library(seasonal)
```

# Task 1

[Notes]{.underline}

```{r Obtain Data (Quarterly)}

api_key <- '3949fd9fcfaa43711339ce7b8a243180'

fredr_set_key(api_key)

# Q1: 01-01, Q2: 04-01, Q3: 07-01, Q4: 10-01

# Federal Funds Effective Rate (FFER)
ffer_full <- fredr(
  series_id = "FEDFUNDS",
  frequency = "q",              # quarterly
  aggregation_method = "avg",
  observation_start = as.Date("1980-01-01")
)

head(ffer_full)



# Personal Consumption Expenditures Index (PCEPI) - Inflation measure
pcepi_full <- fredr(
  series_id = "PCEPI",
  frequency = "q",
  aggregation_method = "avg",
  observation_start = as.Date("1980-01-01")
)

head(pcepi_full)



# Real GDP
gdp_full <- fredr(
  series_id = "GDPC1",
  frequency = "q",
  aggregation_method = "avg",
  observation_start = as.Date("1980-01-01")
)

head(gdp_full)


# Potential Real GDP
pot_gdp_full <- fredr(
  series_id = "GDPPOT",
  frequency = "q",
  aggregation_method = "avg",
  observation_start = as.Date("1980-01-01")
)

head(pot_gdp_full)


```

```{r Construct Variables}
# Add pct_chg_pce col to PCE series
t_val <- pcepi_full$value
t_minus_one_val <- lag(pcepi_full$value, 1)
pcepi_full$pct_chg_pce <- (t_val - t_minus_one_val) / t_minus_one_val


gdp_merged <- inner_join(gdp_full, pot_gdp_full, by='date')

# Add Output Gap col to GDP series
gdp_merged$output_gap <- (gdp_merged$value.x - gdp_merged$value.y) * 100 / gdp_merged$value.y

gdp_merged$series_id <- 'OUTPUTGAP'
gdp_merged <- gdp_merged %>% rename(realtime_start = realtime_start.x, realtime_end = realtime_end.x, value=output_gap)

output_gap <- gdp_merged %>% select(date, series_id, value, realtime_start, realtime_end)
```

```{r FFER Plot + Summary Statistics}

ffer_full %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_line() + 
  labs(title = "Federal Funds Effective Rate (FFER)",
       x = "date",
       y = 'FFER')

ffer_full %>%
  summarize(
    mean = mean(value, na.rm = TRUE),
    var_value  = var(value, na.rm = TRUE),
    sd_value   = sd(value, na.rm = TRUE),
    n_obs      = n()
  )

```

One may note a few key historical events that appear in the graph:

-   A near-zero interest rate between 2008 and 2015. This rate was a result of the 2008 Grand Recession

<!-- -->

-   The 1980s had very high interest rates in order to curb high rates of inflation

-   Rates were additionally cut in 2000 in response to the dot-com bubble

-   In 2020, rates were cut to near-zero as a result of the pandemic

```{r PCEPI Plot + Summary Statistics}
pcepi_full %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_line() + 
  labs(title = "Personal Consumption Expenditures Index (PCEPI)",
       x = "date",
       y = 'PCEPI')

pcepi_full %>%
  summarize(
    mean = mean(value, na.rm = TRUE),
    var_value  = var(value, na.rm = TRUE),
    sd_value   = sd(value, na.rm = TRUE),
    n_obs      = n()
  )

pcepi_full %>% 
  ggplot(aes(x = date, y = pct_chg_pce)) + 
  geom_line() + 
  labs(title = "Pct Change PCE",
       x = "date",
       y = 'Pct Change PCE')
```

PCEPI (Inflation) shows events that are related to events that spurred the FFER changes we see in the above graph.

-   

-   High inflation characterized the early 1980s, as shown by the peak in the graph

-   Very low inflation appears in 2008 during the 2008 financial crisis 

-   Inflation hit local maxima around 2020 as a result of the pandemic

```{r GDP Plot + Summary Statistics}
# Do regular GDP, POT GDP, and output gap

gdp_full %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_line() + 
  labs(title = "Real GDP ",
       x = "date",
       y = 'GDP')

gdp_full %>%
  summarize(
    mean = mean(value, na.rm = TRUE),
    var_value  = var(value, na.rm = TRUE),
    sd_value   = sd(value, na.rm = TRUE),
    n_obs      = n()
  )

pot_gdp_full %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_line() + 
  labs(title = "Potential Real GDP ",
       x = "date",
       y = 'GDPPOT')

pot_gdp_full %>%
  summarize(
    mean = mean(value, na.rm = TRUE),
    var_value  = var(value, na.rm = TRUE),
    sd_value   = sd(value, na.rm = TRUE),
    n_obs      = n()
  )

output_gap %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_line() + 
  labs(title = "Output gap",
       x = "date",
       y = 'Output gap')
```

GDP appears to steadily increase from 1980 to 2020, with a notable dip in 2020 due the COVID pandemic. The output gap timeseries is affected by similar events as FFER and PCEPI:

-   A dip in the early 1980s is seen, which is at the same time as high inflation and spiked interest rates

-   A major dip in 2008 exists, which is at the time of the 2008 financial crisis

-   A sharp dip in 2020 is seen, corresponding to the time of the COVID pandemic

# Task 2

**2.1: Frequency Alignment (Resampling)**

Based on the format of the API call to FRED, data is received at a quarterly cadence, with an average over all months in the quarter calculated and returned in the dataframe. No resampling was done.

```{r Ensure same quarterly date index for all series}


ensure_same_dates <- ffer_full %>% 
  full_join(gdp_full, by='date') %>%
  full_join(pcepi_full, by='date') %>% 
  full_join(pot_gdp_full, by='date') %>%
  full_join(output_gap, by='date')

# Filter all to 2025-04-01 after visual examination of the data
ffer_full <- ffer_full %>% filter(date <= as.Date('2025-04-01'))
pcepi_full <- pcepi_full %>% filter(date <= as.Date('2025-04-01'))
pot_gdp_full <- pot_gdp_full %>% filter(date <= as.Date('2025-04-01'))
output_gap <- output_gap %>% filter(date <= as.Date('2025-04-01'))

```

**2.2: Outlier Detection & Treatment**

[Method: Hampel filter]{.underline}

To detect outliers, we've chosen to use the Hampel filter. We prefer this method because based on a visual examination of FFER, PECPI, and GDP data, scores appear to be mostly following either an upward trend or a fluctuating downward trend, with few outliers. The outliers that do appear (i.e. in Real GDP around 2020) are within the data (as opposed to on the upper and lower ends of the data). Thus, this would not be caught by winsorization. The rolling window calculations from a Hampel filter (relying on deviations from a median) would catch these outliers and smooth out the data. Z-scores, as well, is best suited to normally distributed data, which we don't have.

[Outlier Detection]{.underline}

We utilize a Hampel filter for the FFER, PECPI, GDP, GDPPOT, and output gap timeseries. A window size of 5 appeared to smooth out outliers best upon a visual examination of the corrected timeseries data.

```{r Apply Hampel filters}

ffer_full$hampel_val <- hampel(ffer_full$value, k = 5)[[1]]

ffer_full %>%
  ggplot(aes(x = date)) + 
  geom_line(aes(y = hampel_val, color = "Hampel")) +
  geom_line(aes(y = value, color = "Original")) +
  labs(title = "Federal Funds Effective Rate (FFER)",
       x = "Date",
       y = "Value") +
  scale_color_manual(values = c("Hampel" = "blue", "Original" = "red"),
                     name = "Legend")


```

```{r}
pcepi_full$hampel_val <- hampel(pcepi_full$value, k = 5, )[[1]]

pcepi_full %>%
  ggplot(aes(x = date)) + 
  geom_line(aes(y = hampel_val, color = "Hampel")) +
  geom_line(aes(y = value, color = "Original")) +
  labs(title = "PCEPI",
       x = "Date",
       y = "Value") +
  scale_color_manual(values = c("Hampel" = "blue", "Original" = "red"),
                     name = "Legend")
```

```{r}
gdp_full$hampel_val <- hampel(gdp_full$value, k = 5, )[[1]]

gdp_full %>%
  ggplot(aes(x = date)) + 
  geom_line(aes(y = hampel_val, color = "Hampel")) +
  geom_line(aes(y = value, color = "Original")) +
  labs(title = "GDP",
       x = "Date",
       y = "Value") +
  scale_color_manual(values = c("Hampel" = "blue", "Original" = "red"),
                     name = "Legend")
```

```{r}
pot_gdp_full$hampel_val <- hampel(pot_gdp_full$value, k = 5, )[[1]]

pot_gdp_full %>%
  ggplot(aes(x = date)) + 
  geom_line(aes(y = hampel_val, color = "Hampel")) +
  geom_line(aes(y = value, color = "Original")) +
  labs(title = "GDPPOT",
       x = "Date",
       y = "Value") +
  scale_color_manual(values = c("Hampel" = "blue", "Original" = "red"),
                     name = "Legend")
```

```{r}
output_gap$value <- (gdp_full$hampel_val - pot_gdp_full$hampel_val) * 100 / pot_gdp_full$hampel_val

```

**2.3: Seasonal Adjustment**

[Detection]{.underline}

```{r FFER Seasonal Plots + QS Test}

ffer_full <- ffer_full %>% mutate(q = quarter(date))


ffer_full$q_factor <- factor(ffer_full$q, levels = c("1", "2", "3", "4"))

ggplot(ffer_full, aes(x = q_factor, y = value)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(title = "Distribution of Values by Quarter",
       x = "Quarter",
       y = "Value") +
  theme_minimal()

ffer_full %>% filter(q == 1) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "FFER Q1", x = "Date", y = "Value")

ffer_full %>% filter(q == 2) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "FFER Q2", x = "Date", y = "Value")

ffer_full %>% filter(q == 3) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "FFER Q3", x = "Date", y = "Value")

ffer_full %>% filter(q == 4) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "FFER Q4", x = "Date", y = "Value")

start_year <- 1980
start_quarter <- 1

ffer_ts <- ts(ffer_full$hampel_val, start = c(start_year, start_quarter), frequency=4)

qs(ffer_ts)
```

According to our QS test, no evidence for seasonality in the FFER data exists

```{r PCEPI Seasonal Plots + QS Test}
pcepi_full <- pcepi_full %>% mutate(q = quarter(date))

pcepi_full$q_factor <- factor(pcepi_full$q, levels = c("1", "2", "3", "4"))

ggplot(pcepi_full, aes(x = q_factor, y = value)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(title = "Distribution of Values by Quarter",
       x = "Quarter",
       y = "Value") +
  theme_minimal()

pcepi_full %>% filter(q == 1) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "PCEPI Q1", x = "Date", y = "Value")

pcepi_full %>% filter(q == 2) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "PCEPI Q2", x = "Date", y = "Value")

pcepi_full %>% filter(q == 3) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "PCEPI Q3", x = "Date", y = "Value")

pcepi_full %>% filter(q == 4) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "PCEPI Q4", x = "Date", y = "Value")

start_year <- 1980
start_quarter <- 1

pcepi_ts <- ts(pcepi_full$hampel_val, start = c(start_year, start_quarter), frequency=4)

qs(pcepi_ts)
```

According to our QS test, evidence for seasonality in the PCEPI data exists.

```{r GDP Seasonal Plots + QS Test}
gdp_full <- gdp_full %>% mutate(q = quarter(date))

gdp_full$q_factor <- factor(gdp_full$q, levels = c("1", "2", "3", "4"))

ggplot(gdp_full, aes(x = q_factor, y = value)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(title = "Distribution of Values by Quarter",
       x = "Quarter",
       y = "Value") +
  theme_minimal()

gdp_full %>% filter(q == 1) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "GDP Q1", x = "Date", y = "Value")

gdp_full %>% filter(q == 2) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "GDP Q2", x = "Date", y = "Value")

gdp_full %>% filter(q == 3) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "GDP Q3", x = "Date", y = "Value")

gdp_full %>% filter(q == 4) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "GDP Q4", x = "Date", y = "Value")

start_year <- 1980
start_quarter <- 1

gdp_full_ts <- ts(gdp_full$hampel_val, start = c(start_year, start_quarter), frequency=4)

qs(gdp_full_ts)
```

According to our QS test, no evidence for seasonality in the GDP data exists

```{r GDPPOT Seasonal Plots + QS Test}
pot_gdp_full <- pot_gdp_full %>% mutate(q = quarter(date))

pot_gdp_full$q_factor <- factor(pot_gdp_full$q, levels = c("1", "2", "3", "4"))

ggplot(pot_gdp_full, aes(x = q_factor, y = value)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(title = "Distribution of Values by Quarter",
       x = "Quarter",
       y = "Value") +
  theme_minimal()

pot_gdp_full %>% filter(q == 1) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "GDPPOT Q1", x = "Date", y = "Value")

pot_gdp_full %>% filter(q == 2) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "GDPPOT Q2", x = "Date", y = "Value")

pot_gdp_full %>% filter(q == 3) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "GDPPOT Q3", x = "Date", y = "Value")

pot_gdp_full %>% filter(q == 4) %>% ggplot(aes(x=date, y=hampel_val)) + geom_line() + labs(title = "GDPPOT Q4", x = "Date", y = "Value")

start_year <- 1980
start_quarter <- 1

pot_gdp_full_ts <- ts(pot_gdp_full$hampel_val, start = c(start_year, start_quarter), frequency=4)

qs(pot_gdp_full_ts)
```

According to our QS test, evidence for seasonality in the GDPPOT data exists

[Seasonal Adjustment via]{.underline} **X-13ARIMA-SEATS**

This method is used because it allows for timeseries that can be decomposed additively or multiplicatively, allowing for maximum flexibility. The trend, seasonal component, and irregular component are all estimated and the algorithm uses centered moving averages to remove the seasonal component.

Source: <https://en.wikipedia.org/wiki/X-13ARIMA-SEATS>

```{r}

pcepi_fit <- seas(pcepi_ts)

pcepi_full$s_adj_val <- final(pcepi_fit)

pot_gdp_fit <- seas(pot_gdp_full_ts)

pot_gdp_full$s_adj_val <- final(pot_gdp_fit)


pot_gdp_full %>%
  ggplot(aes(x = date)) + 
  geom_line(aes(y = s_adj_val, color = "Seasonally Adjusted")) +
  geom_line(aes(y = value, color = "Original")) +
  labs(title = "GDPPOT",
       x = "Date",
       y = "Value") +
  scale_color_manual(values = c("Seasonally Adjusted" = "blue", "Original" = "red"),
                     name = "Legend")

pcepi_full %>%
  ggplot(aes(x = date)) + 
  geom_line(aes(y = s_adj_val, color = "Seasonally Adjusted")) +
  geom_line(aes(y = value, color = "Original")) +
  labs(title = "PCEPI",
       x = "Date",
       y = "Value") +
  scale_color_manual(values = c("Seasonally Adjusted" = "blue", "Original" = "red"),
                     name = "Legend")
```

```{r}
# # Split into train and test set 
# ffer_train <- ffer_full %>% filter(date >= as.Date('1980-01-01') & date <= as.Date('2023-10-01')) 

# ffer_test <- ffer_full %>% filter(date >= as.Date('2024-01-01')) 

#  # # Split into train and test set 

# pcepi_train <- pcepi_full %>% filter(date >= as.Date('1980-01-01') & date <= as.Date('2023-10-01')) 

# pcepi_test <- pcepi_full %>% filter(date >= as.Date('2024-01-01')) 

#  # # Split into train and test set 

# gdp_train <- gdp_full %>% filter(date >= as.Date('1980-01-01') & date <= as.Date('2023-10-01')) 

# gdp_test <- gdp_full %>% filter(date >= as.Date('2024-01-01')) 

#  # # Split into train and test set 
# pot_gdp_train <- pot_gdp_full %>% filter(date >= as.Date('1980-01-01') & date <= as.Date('2023-10-01')) 

# pot_gdp_test <- pot_gdp_full %>% filter(date >= as.Date('2024-01-01'))
```
